\thetitle

### 2008-06-26 paper.tex
Most of John's comments.

\begin{abstract}
We propose a representation of wireless workload patterns as large, sparse matrices and provide a method for stochastically generating experimental workload from a given matrix. The essential property of the algebraic representation is that the summation of vectors naturally yields a faithful description of the aggregate behavior of the corresponding flows. This deceptively simple property allows us to express many common concepts from traffic modeling succinctly in terms of a few linear transformations. The algebraic representation has many benefits: 1)~it makes the meaning of generally understood but vague concepts, such as ``uniform behavior,'' mathematically precise and unambiguous; 2)~it allows us to see clearly, through the lens of linear algebra, the implications of common modeling assumptions; 3)~the implementation of traffic models becomes unprecedentedly simple and orthogonal, requiring only a handful of high-level matrix operations, which can be freely composed; 4)~the vast body of algebraic theory and highly optimized numerical linear software may immediately be applied to traffic modeling. We use the paired differential simulation methodology introduced by the authors in previous work to experimentally demonstrate that the general matrix model accurately reproduces realistic network performance~\cite{Karpinski07:realism,Karpinski07:cbr-failure}. We use the same experimental methodology to explore the implications of various assumptions and simplifications that are commonly made in traffic modeling.
\end{abstract}

@@

More than twenty years of research in analysis and modeling of network traffic have yielded tremendous advances in our understanding of the complex behaviors that emerge from the interaction of millions of humans and computers on the Internet and in local-area networks (\caps{LAN}s).
Both traffic analysis and realistic workload generation, however, remain active areas of research with many unanswered questions.
One of the major challenges of modeling whole-network traffic is the interdependence of packet, flow\flownote and node behaviors.
Three types of simplifying assumptions are commonly applied in both analysis and generation:
\begin{enumerate}
\item \textbf{Uniform:} that traffic properties are uniformly distributed. For example, assuming that  all nodes in a network have equal probability of being the source or destination of each flow.
Another common example is the assumption that all packet sizes and inter-packet intervals are the same; this is commonly known as the ``constant bit-rate'' (\caps{CBR}) model.
\item \textbf{Marginal:} that traffic properties share a single, common \emph{marginal}\marginalnote distribution across the entire network, and individual values are drawn independently and unconditionally from this shared distribution.
For example, assuming that a single distribution of packet sizes describes all flows in a network, as compared to equipping each flow with its own specific distribution of sizes.
\item \textbf{Conditional:} that traffic behaves marginally within equivalence classes defined by certain conditions.
The assumption, for example, that all flows emanating from a common source node share the same packet size distribution is a conditional model, conditioned on the source nodes.
\end{enumerate}
We would be arguing against a straw man if we implied that researchers who apply these assumptions actually believe them to be true.
They clearly are not:
different nodes initiate and receive drastically different numbers of packets and flows;
flows have drastically different packet size distributions.
Simply considering intuitive examples of common behavior illustrates this amply:
BitTorrent users \latin{vs.} occasional email checkers;
downloading a large file \latin{vs.} typing in an \caps{SSH} session.

@@

Once traffic is represented linearly, alternatives to assuming uniformity or marginality become apparent.
We argue that matrix factorization, which is nondestructive and does not assume uniformity or marginality, is far superior to multiplication for this problem.
To demonstrate this approach, we use nonnegative matrix factorization (\caps{NMF})~\cite{Lee01} to find a low-dimensional approximation of the packet size and inter-packet interval matrices of a real traffic trace.
This factorization extracts a small set of ``basic behaviors'' from the traffic matrix and simultaneously explains each flow's observed behavior as a mixture of these basic behaviors.
Although the factorization is computed without examining port numbers, basic behaviors nevertheless correspond to intuitive expectations for protocols.
We find basic behaviors corresponding to typical network activities, including file transfer, typing via \caps{SSH}, web surfing, ping traffic, peer-to-peer and others.

@@

The rest of the paper is organized as follows. In \Section{motivation} we discuss motivation and related work. The notion of linear representation and the general matrix model are presented in \Section{general-matrix-model}. In Sections~\ref{sec:common-properties}~and~\ref{sec:modeling-concepts} we explore how common network properties and modeling concepts can be expressed in the the \caps{GMM} framework. Our experimental and analytical methodology for evaluating traffic models is presented in \Section{methodology}, while the results of our experiments are explained and analyzed in \Section{results}. The ramifications of these results are discussed in \Section{discussion}. Finally, in \Section{conclusions}, we conclude with a discussion of how this research may be applied to current wireless studies.

@@

The interaction of network users and applications with the lower layers of the networking stack is characterized by where, when, how much, and to whom data is transmitted. The joint pattern of traffic generation and mobility through time and space completely determines the effect of network usage on the lower levels of the stack. This is due to the data-agnostic nature of the protocol stack: by design, \caps{IP} networks treat all data in the same manner~\cite{Clark88}.\QoSnote The credibility of conclusions derived from simulation or experimental deployment depends crucially on our confidence that the models used to generate traffic in experiments are sufficiently realistic. \Figure{trace-vs-uniform} illustrates through examples how drastically different network traffic appears under assumptions of uniformity at different levels of behavior. Marginality assumptions are not as visually striking, but as we will demonstrate, they distort the performance characteristics of networks almost as much, and sometimes more, than uniform models.

@@

Paxson and Floyd observed~\cite{Paxson95} that the interaction between endpoint behavior and the network conditions is inherently \textit{closed-loop} in the sense that the emergent behavior is governed by a closed feedback loop.
In the case of most non-\caps{TCP} applications, this feedback does not occur: an open-loop model suffices.
Hern\'andez-Campos' dissertation~\cite{Hernandez06:dissertation} addresses in great depth how to model the feedback between the network and typical \caps{TCP} applications.
The author shows how to abstract an application's behavior from a \caps{TCP} trace and replay it in a manner that accurately reflects how typical applications react in response to different network conditions.
This research has subsequently been turned into a \caps{TCP} flow generation tool called \tmix~\cite{Weigle06}.
The acronym is somewhat misleading: {\tmix} does not provide models for application behaviors or for mixtures of applications in networks.
Rather, it provides the ability to reproduce accurate closed-loop dynamics for a single \caps{TCP} flow \emph{given} an application behavior model as input.
Our research complements this perfectly: we provide the means to analyze traffic and generate realistic whole-network application mixtures which can then serve as the necessary inputs to {\tmix}.

@@

Like Harpoon, \caps{D-ITG}~\cite{Avallone04} exemplifies uniform and marginal modeling assumptions. However, unlike Harpoon, \caps{D-ITG} does not derive models from real traffic.
Packet behavior can be modeled using standard statistical distributions (``constant, uniform, normal, cauchy, pareto, exponential, etc.'').
Guidance is not provided, however, for choosing among these distributions, choosing parameter values, or selecting a mixture of distributions across a network.
Flow and node behaviors are not specified at all, but rather left to the user's discretion.
The focus of this project is primarily on the engineering challenge of generating very large volumes of synthetic traffic and injecting it into networks in a distributed fashion.
While \caps{D-ITG} provides the mechanism for generating large volumes of traffic, it does not provide any real guidance as to what traffic to generate.

@@

Our first example is the packets-bytes-duration (\caps{PBD}) representation.
The representation is $\pbd: \Traffic \to \R^3$, mapping
$\flow \mapsto \tuple{p,b,d} \in \R^3$, where
\begin{align}
p &= \pkts, &
b &= \sum_{i=1}^{\mathclap{p}}{\size_i}, &
d &= \sum_{i=1}^{\mathclap{p-1}}{\ival_i}.
\end{align}
This representation encapsulates the information necessary to reproduce a \caps{CBR} version of each flow:
packets have payloads of $\floor{\fracx{b}{p}}$ bytes, and are sent every $\fracx{d}{p}$ seconds.\pbdnote
Given two flows with behavior vectors $\vec{x}_1$ and $\vec{x}_2$, their sum as vectors, 
\begin{align}
\vec{x}_1+\vec{x}_2=\tuple{p_1+p_2,b_1+b_2,d_1+d_2},
\end{align}
provides an appropriate description of the aggregate behavior of the two flows together.
Similarly, their average, \latin{i.e.} $\smallfrac{1}{2}(\vec{x}_1+\vec{x}_2)$, has the average number of packets, bytes, and duration, which is precisely what the ``average behavior'' of the two flows should intuitively be.
In contrast, a ``packets-payload-interval'' representation, expressing the \caps{CBR} parameters directly, does not have this property:
the sum of average payloads is not the overall average of payloads;
likewise for intervals.
In the \caps{PBD} representation, the linearity property leads to natural descriptions of aggregate behavior through vector addition, whereas in the packets-payload-interval representation, vector addition is ``unnatural.''

@@

The \caps{PBD} representation of flows uses 3 dimensions to encode the behavior of each flow, while $\Phi$ uses $255+2n$ dimensions, plus the continuum of dimensions of two Banach spaces (\latin{i.e.} the two copies of $L^2[\R]$).
In this section, we define a representation, $\phi$, that splits the difference and uses a large but finite number of dimensions to represent each flow.
Once the vector representation of each flow is defined, we may construct matrices of representation vectors, with a row for each flow in a traffic sample.
We call the matrix representation of network traffic the \emph{general matrix model}~(\caps{GMM}).
Having expressed network behavior in terms of matrices, we can very simply and efficiently compute a broad variety of network behavior properties using standard matrix operations.
Furthermore, we can express uniformity and marginality assumptions about behavior as equality constraints between the rows and columns of the traffic matrix.
Moreover, we can transform unconstrained traffic instances into uniform, marginal, or conditional approximations via right or left multiplications by easily computed matrices.

@@

The quantization functions specified in \Table{quantization} allow us to define the representation $\phi: \Traffic \to \R^N$: \begin{align}
\phi(\flow) = \vec{f} = \tuple{\vec{t},\vec{s},\vec{d},\vec{a},\vec{b},\vec{z},\vec{v}},
\end{align}
where $N=255+2n+d_t+d_b+d_z+d_v$ and vectors are
\begin{align*}
\vec{t} &= \e_k & k &= \Qf_\type(\type) \\
\vec{s} &= \e_k & k &= \Qf_\src(\src) \\
\vec{d} &= \e_k & k &= \Qf_\dst(\dst) \\
\vec{a} &= \e_k & k &= \Qf_\start(\start) \\
\vec{b} &= \e_k & k &= \Qf_\bytes(\bytes) \\
\vec{z} &= {\textstyle \sum \e_k} & k &\in \set{\Qf_\size(\size_i) : i \le \pkts} \\
\vec{v} &= {\textstyle \sum \e_k} & k &\in \set{\Qf_\ival(\ival_i) : i <   \pkts}.
\end{align*}
The functions $\Qf_\prop$ are the respective quantization functions for each property.
If there are $f$ flows in a traffic instance, then the collective behavior can be described with seven matrices---one for each property;
the rows of each matrix are the behavior vectors of the flows, indexed in some chosen order.
We use the following notations for the matrices and indexed flow vectors:
\vspace{-0.25em}
\begin{tabbing}
\hspace{1em}\=\bullet\hspace{0.5em}\=\hspace{9.75em}\=
$\mat{D}\:$\=$=\parensx{\vec{d}_i}\:$\=$=\parensx{d_{ij}}\:$\=\kill
\>\bullet\> types: \>$\mat{T}$\>$=\parensx{\vec{t}_i}$\>$=\parensx{t_{ij}}$\>$\in\R^{f \times 255}$\\
\>\bullet\> sources: \>$\mat{S}$\>$=\parensx{\vec{s}_i}$\>$=\parensx{s_{ij}}$\>$\in\R^{f \times n}$\\
\>\bullet\> destinations: \>$\mat{D}$\>$=\parensx{\vec{d}_i}$\>$=\parensx{d_{ij}}$\>$\in\R^{f \times n}$\\
\>\bullet\> start times: \>$\mat{A}$\>$=\parensx{\vec{a}_i}$\>$=\parensx{a_{ij}}$\>$\in\R^{f \times d_t}$\\
\>\bullet\> sizes in bytes: \>$\mat{B}$\>$=\parensx{\vec{b}_i}$\>$=\parensx{b_{ij}}$\>$\in\R^{f \times d_b}$\\
\>\bullet\> packet sizes: \>$\mat{Z}$\>$=\parensx{\vec{z}_i}$\>$=\parensx{z_{ij}}$\>$\in\R^{f \times d_z}$\\
\>\bullet\> inter-packet intervals: \>$\mat{V}$\>$=\parensx{\vec{v}_i}$\>$=\parensx{v_{ij}}$\>$\in\R^{f \times d_v}$.
\end{tabbing}
The total behavior matrix is the horizontal concatenation of these component matrices:
\begin{align}
\mat{F} = \left[\,\mat{T}~\mat{S}~\mat{D}~\mat{A}~\mat{B}~\mat{Z}~\mat{V}\,\right] \in \R^{f \times N}.
\end{align}
We call this total representation the \emph{general matrix model}.
While it is by no means the only possible representation of traffic in terms of matrices, as we demonstrate in the following sections, many common traffic modeling assumptions can be expressed succinctly and precisely in terms of transformations or conditions of the \caps{GMM} and its components.
Furthermore, realistic workload can be generated from matrix instances.

@@

Before continuing with less mundane topics, we must briefly address our choices of constants appearing in \Table{quantization}.
The externally predetermined constants are:
$d_z=\text{\caps{MTU}}=1500$, determined by the wireless network medium;
$b_{\max} \approx 109 \text{\caps{MB}}$, determined by the size of the largest flow in our trace data;
$t_{\max}=v_{\max}=600$, determined by our experimental methodology which slices traces into 10-minute scenarios;
and $v_{\min}=10^{-6}$, determined by the time resolution of our trace.
For lack of a better choice, we have simply chosen to use the same number of quantization bins as packet size for the unfixed dimension constants:
$d_t = d_b = d_v = 1500$.
Arguments can certainly be made for other choices, but this choice appears to work adequately in practice.

@@

From the matrices $\mat{Z}$ and $\mat{V}$ we can compute a variety of statistics about packet sizes and inter-packet intervals.
The flows vectors in the matrices may represent individual flows or composites of many flows.
% about packet sizes, total packet count, inter-packet intervals, and total duration for each representation vector, which may represent a single actual flow, or the composite behavior of many flows.
% We begin by expressing the vectors for total bytes, total packets and total duration for a matrix of flow representations, with size and interval matrices $\mat{Z}$ and $\mat{V}$.
The computations require knowing the values that each size of interval matrix column represents.
% The computed duration is approximate because information is irretrievably lost in the quantization process.
The vectors of representative values for $\mat{Z}$ and $\mat{V}$ respectively are:
\begin{align}
\vec{\sigma} &= \tuple{1,2,3,\dots,d_z} \\
\vec{\tau}   &= \tuple{\Df_\ival\left(k-\smallfrac{1}{2}\right)}_{k=1}^{d_v}.
\end{align}
The packet count, total byte and total duration vectors are computed as
\begin{align}
\vec{\Sigma}_\pkts  &= \mat{Z}\trans{\ones} = \mat{V}\trans{\ones} \in \R^f \eqn{packet-count}\\
\vec{\Sigma}_\bytes &= \mat{Z}\trans{\vec{\sigma}} \in \R^f \eqn{total-bytes} \\
\vec{\Sigma}_\ivals &= \mat{V}\trans{\vec{\tau}} \in \R^f \eqn{total-duration}.
\end{align}
Here $\ones$ is a row vector of all ones.\onesnote
Packet count and total bytes values are exact;
the total durations are only approximate because information is irretrievably lost in the quantization process.
Using these vectors of totals, we may readily compute the average packet sizes and inter-packet intervals:
\begin{align}
\vec{\mean{z}} &=
  \vec{\Sigma}_\bytes \eldiv \vec{\Sigma}_\pkts
  \phantom{= \mat{V}\trans{\vec{\tau}} \eldiv \mat{V}\trans{\ones}}
  \mathllap{= \mat{Z}\trans{\vec{\sigma}} \eldiv \mat{Z}\trans{\ones}} \\
\vec{\mean{v}} &=
  \phantom{\vec{\Sigma}_\bytes \eldiv \vec{\Sigma}_\pkts}
  \mathllap{\vec{\Sigma_{\ivals}} \eldiv \vec{\Sigma_{\pkts}}}
  = \mat{V}\trans{\vec{\tau}} \eldiv \mat{V}\trans{\ones}.
\end{align}
The symbol $\eldiv$ here represents element-wise division of vectors, as it does in Matlab.

@@

Given an instance of the general matrix model, how do we produce an actual workload from it?
Each row of the model is used to randomly generate the behavior of a single flow.
To do this, we use the ``dequantization'' functions listed in \Table{quantization}.
The dequantization procedure works as follows.
Let $\vec{f}$
% \begin{align}
% \vec{f}=\tuple{\vec{t},\vec{s},\vec{d},\vec{a},\vec{b},\vec{z},\vec{v}}
% \end{align}
be a row of the traffic matrix, and let $\prop$ be a property.
Write $\vec{f}_\prop$ for the corresponding component vector of $\vec{f}$.
Chose an index value $k\in\set{1,\dots,d_\prop}$, randomly, with weight given by the coefficients of $\vec{f}_\prop$.
Choose $u \in [0,1]$ uniformly randomly.
Then $x = k-u \in [0,d]$, with probability of lying in the interval $[i-1,i]$ proportional to the coefficient of $\e_i$ in $\vec{f}_\prop$.
The dequantization function for $\prop$, namely $\Df_\prop$, will map the value $x$ back into the appropriate range of property values for $\prop$.

>>>>

\begin{abstract}
We propose a representation of wireless workload patterns as large, sparse matrices and provide a method for stochastically generating experimental workloads from a given matrix. The essential property of the algebraic representation is that the summation of vectors naturally yields a faithful description of the aggregate behavior of the corresponding flows. This deceptively simple property allows us to express many common concepts from traffic modeling succinctly in terms of a few linear transformations. The algebraic representation has many benefits: 1)~it makes the meaning of generally understood but vague concepts, such as ``uniform behavior,'' mathematically precise and unambiguous; 2)~it allows us to see clearly, through the lens of linear algebra, the implications of common modeling assumptions; 3)~the implementation of traffic models becomes unprecedentedly simple and orthogonal, requiring only a handful of high-level matrix operations, which can be freely composed; 4)~the vast body of algebraic theory and highly optimized numerical  software may immediately be applied to traffic modeling. We use the paired differential simulation methodology introduced by the authors in previous work to experimentally demonstrate that the general matrix model accurately reproduces realistic network performance~\cite{Karpinski07:realism,Karpinski07:cbr-failure}. We use the same experimental methodology to explore the implications of various assumptions and simplifications that are commonly made in traffic modeling.
\end{abstract}

@@

More than twenty years of research in analysis and modeling of network traffic have yielded tremendous advances in our understanding of the complex behaviors that emerge from the interaction of millions of humans and computers on the Internet and in local-area networks (\caps{LAN}s).
Both traffic analysis and realistic workload generation, however, remain active areas of research with many unanswered questions.
One of the major challenges of modeling whole-network traffic is the interdependence of packet, flow,\flownote and node behaviors.
Three types of simplifying assumptions are commonly applied in both analysis and generation:
\begin{enumerate}
\item \textbf{Uniform:} that traffic properties are uniformly distributed. For example, assuming that  all nodes in a network have equal probability of being the source or destination of each flow.
Another common example is the assumption that all packet sizes and inter-packet intervals are the same; this is commonly known as the ``constant bit-rate'' (\caps{CBR}) model.
\item \textbf{Marginal:} that traffic properties share a single, common \emph{marginal}\marginalnote distribution across the entire network, and individual values are drawn independently and unconditionally from this shared distribution.
An example is assuming that a single distribution of packet sizes describes all flows in a network, as compared to equipping each flow with its own specific distribution of sizes.
\item \textbf{Conditional:} that traffic behaves marginally within equivalence classes defined by certain conditions.
The assumption, for example, that all flows emanating from a common source node share the same packet size distribution is a conditional model, conditioned on the source nodes.
\end{enumerate}
We would be arguing against a straw man if we implied that researchers who apply these assumptions actually believe them to be true.
They clearly are not:
different nodes initiate and receive drastically different numbers of packets and flows;
flows have drastically different packet size distributions.
Simply considering intuitive examples of common behavior illustrates this amply:
BitTorrent users \latin{vs.} occasional email checkers;
downloading a large file \latin{vs.} typing in an \caps{SSH} session.

@@

Once traffic is represented linearly, alternatives to assuming uniformity or marginality become apparent.
We argue that matrix factorization, which is nondestructive and does not assume uniformity or marginality, is far superior to multiplication for this problem.
To demonstrate this approach, we use nonnegative matrix factorization (\caps{NMF})~\cite{Lee01} to find a low-dimensional approximation of the packet size and inter-packet interval matrices of a real traffic trace.
This factorization extracts a small set of ``basic behaviors'' from the traffic matrix and simultaneously explains each flow's observed behavior as a mixture of these basic behaviors.
Although the factorization is computed without examining port numbers, basic behaviors nevertheless correspond to intuitive expectations for protocols.
We find basic behaviors corresponding to typical network activities, including file transfer, typing via \caps{SSH}, web surfing, ping traffic, peer-to-peer, and others.

@@

The rest of the paper is organized as follows.
In \Section{motivation} we discuss motivation and related work.
The notion of linear representation is presented in \Section{representations} and the general matrix model is presented in \Section{general-matrix-model}.
Our experimental and analytical methodology for evaluating traffic models is presented in \Section{methodology}, while the results of our experiments are explained and analyzed in \Section{results}.
The ramifications of these results are discussed in \Section{discussion}.
Finally, in \Section{conclusions}, we conclude with a summary of this research and its impact on the future of wireless networking.

@@

The interaction of network users and applications with the lower layers of the networking stack is characterized by where, when, how much, and to whom data is transmitted. The joint pattern of traffic generation and mobility through time and space completely determines the effect of network usage on the lower levels of the stack. This is due to the data-agnostic nature of the protocol stack: by design, \caps{IP} networks treat all data in the same manner~\cite{Clark88}.\QoSnote The credibility of conclusions derived from simulation or experimental deployment depends crucially on our confidence that the models used to generate traffic in experiments are sufficiently realistic. \Figure{trace-vs-uniform} illustrates through examples how drastically different network traffic appears under assumptions of uniformity at different levels of behavior. These examples are take from a wireless trace recorded at an \caps{IETF} meeting; details of the trace are given in \Section{trace-data}. Marginality assumptions are not as visually striking, but as we will demonstrate, they distort the performance characteristics of networks almost as much, and sometimes more, than uniform models.

@@

Paxson and Floyd observed~\cite{Paxson95} that the interaction between endpoint behavior and the network conditions is inherently \textit{closed-loop} in the sense that the emergent behavior is governed by a closed feedback loop.
In the case of most non-\caps{TCP} applications, this feedback does not occur: an open-loop model suffices.
Hern\'andez-Campos's dissertation~\cite{Hernandez06:dissertation} addresses in great depth how to model the feedback between the network and typical \caps{TCP} applications.
The author shows how to abstract an application's behavior from a \caps{TCP} trace and replay it in a manner that accurately reflects how typical applications react in response to different network conditions.
This research has subsequently been turned into a \caps{TCP} flow generation tool called \tmix~\cite{Weigle06}.
The acronym is somewhat misleading: {\tmix} does not provide models for application behaviors or for mixtures of applications in networks.
Rather, it provides the ability to reproduce accurate closed-loop dynamics for a single \caps{TCP} flow \emph{given} an application behavior model as input.
Our research complements this perfectly: we provide the means to analyze traffic and generate realistic whole-network application mixtures which can then serve as the necessary inputs to {\tmix}.

@@

Like Harpoon, \caps{D-ITG}~\cite{Avallone04} exemplifies uniform and marginal modeling assumptions. However, unlike Harpoon, \caps{D-ITG} does not derive models from real traffic.
Packet behavior can be modeled using standard statistical distributions (``constant, uniform, normal, Cauchy, Pareto, exponential, etc.'').
Guidance is not provided, however, for choosing among these distributions, choosing parameter values, or selecting a mixture of distributions across a network.
Flow and node behaviors are not specified at all, but rather left to the user's discretion.
The focus of this project is primarily on the engineering challenge of generating very large volumes of synthetic traffic and injecting it into networks in a distributed fashion.
While \caps{D-ITG} provides the mechanism for generating large volumes of traffic, it does not provide any real guidance as to what traffic to generate.

@@

Our first example is the packets-bytes-duration (\caps{PBD}) representation.
The representation is $\pbd: \Traffic \to \R^3$, mapping
$\flow \mapsto \tuple{p,b,d} \in \R^3$, where
\begin{align}
p &= \pkts, &
b &= \sum_{i=1}^{\mathclap{p}}{\size_i}, &
d &= \sum_{i=1}^{\mathclap{p-1}}{\ival_i}.
\end{align}
This representation encapsulates the information necessary to reproduce a \caps{CBR} version of each flow:
packets have payloads of $\floor{\fracx{b}{p}}$ bytes, and are sent every $\fracx{d}{p}$ seconds.\pbdnote
Given two flows with behavior vectors $\vec{x}_1$ and $\vec{x}_2$, their sum as vectors, 
\begin{align}
\vec{x}_1+\vec{x}_2=\tuple{p_1+p_2,b_1+b_2,d_1+d_2},
\end{align}
provides an appropriate description of the aggregate behavior of the two flows together.
Similarly, their average, $\smallfrac{1}{2}(\vec{x}_1+\vec{x}_2)$, has the average number of packets, bytes, and duration, which is precisely what the ``average behavior'' of the two flows should intuitively be.
In contrast, a ``packets-payload-interval'' representation, expressing the \caps{CBR} parameters directly, does not have this property:
the sum of average payloads is not the overall average of payloads;
likewise for intervals.
In the \caps{PBD} representation, the linearity property leads to natural descriptions of aggregate behavior through vector addition, whereas in the packets-payload-interval representation, vector addition is ``unnatural.''

@@

The \caps{PBD} representation of flows uses three dimensions to encode the behavior of each flow, while $\Phi$ uses $255+2n$ dimensions, plus the continuum of dimensions of two Banach spaces (\latin{i.e.} the two copies of $L^2[\R]$).
In this section, we define a representation, $\phi$, that splits the difference and uses a large but finite number of dimensions to represent each flow.
Once the vector representation of each flow is defined, we may construct matrices of representation vectors, with a row for each flow in a traffic sample.
We call the matrix representation of network traffic the \emph{general matrix model}~(\caps{GMM}).
Having expressed network behavior in terms of matrices, we can very simply and efficiently compute a broad variety of network behavior properties using standard matrix operations.
Furthermore, we can express uniformity and marginality assumptions about behavior as equality constraints between the rows and columns of the traffic matrix.
Moreover, we can transform unconstrained traffic instances into uniform, marginal, or conditional approximations via right or left multiplications by easily computed matrices.

@@

\subsection{Model Specification}

The quantization functions specified in \Table{quantization} allow us to define the representation $\phi: \Traffic \to \R^N$: \begin{align}
\phi(\flow) = \vec{f} = \tuple{\vec{t},\vec{s},\vec{d},\vec{a},\vec{b},\vec{z},\vec{v}},
\end{align}
where $N=255+2n+d_a+d_b+d_z+d_v$ and vectors are
\begin{align*}
\vec{t} &= \e_k & k &= \Qf_\type(\type) \\
\vec{s} &= \e_k & k &= \Qf_\src(\src) \\
\vec{d} &= \e_k & k &= \Qf_\dst(\dst)
\end{align*}
\begin{align*}
\vec{a} &= \e_k & k &= \Qf_\start(\start) \\
\vec{b} &= \e_k & k &= \Qf_\bytes(\bytes) \\
\vec{z} &= {\textstyle \sum \e_k} & k &\in \set{\Qf_\size(\size_i) : i \le \pkts} \\
\vec{v} &= {\textstyle \sum \e_k} & k &\in \set{\Qf_\ival(\ival_i) : i <   \pkts}.
\end{align*}
The functions $\Qf_\prop$ are the respective quantization functions for each property.
If there are $f$ flows in a traffic instance, then the collective behavior can be described with seven matrices---one for each property;
the rows of each matrix are the behavior vectors of the flows, indexed in some chosen order.
We use the following notations for the matrices and indexed flow vectors:
\vspace{-0.25em}
\begin{tabbing}
\hspace{1em}\=\bullet\hspace{0.5em}\=\hspace{9.75em}\=
$\mat{D}\:$\=$=\parensx{\vec{d}_i}\:$\=$=\parensx{d_{ij}}\:$\=\kill
\>\bullet\> types: \>$\mat{T}$\>$=\parensx{\vec{t}_i}$\>$=\parensx{t_{ij}}$\>$\in\R^{f \times 255}$\\
\>\bullet\> sources: \>$\mat{S}$\>$=\parensx{\vec{s}_i}$\>$=\parensx{s_{ij}}$\>$\in\R^{f \times n}$\\
\>\bullet\> destinations: \>$\mat{D}$\>$=\parensx{\vec{d}_i}$\>$=\parensx{d_{ij}}$\>$\in\R^{f \times n}$\\
\>\bullet\> start times: \>$\mat{A}$\>$=\parensx{\vec{a}_i}$\>$=\parensx{a_{ij}}$\>$\in\R^{f \times d_a}$\\
\>\bullet\> sizes in bytes: \>$\mat{B}$\>$=\parensx{\vec{b}_i}$\>$=\parensx{b_{ij}}$\>$\in\R^{f \times d_b}$\\
\>\bullet\> packet sizes: \>$\mat{Z}$\>$=\parensx{\vec{z}_i}$\>$=\parensx{z_{ij}}$\>$\in\R^{f \times d_z}$\\
\>\bullet\> inter-packet intervals: \>$\mat{V}$\>$=\parensx{\vec{v}_i}$\>$=\parensx{v_{ij}}$\>$\in\R^{f \times d_v}$.
\end{tabbing}
The total behavior matrix is the horizontal concatenation of these component matrices:
\begin{align}
\mat{F} = \left[\,\mat{T}~\mat{S}~\mat{D}~\mat{A}~\mat{B}~\mat{Z}~\mat{V}\,\right] \in \R^{f \times N}.
\end{align}
We call this total representation the \emph{general matrix model}.
While it is by no means the only possible representation of traffic in terms of matrices, as we demonstrate in the following sections, many common traffic modeling assumptions can be expressed succinctly and precisely in terms of transformations or conditions of the \caps{GMM} and its components.
Furthermore, realistic workload can be generated from matrix instances.

@@

Before continuing with less mundane topics, we must briefly address our choices of constants appearing in \Table{quantization}.
The externally predetermined constants are:
$d_z=\text{\caps{MTU}}=1500$, determined by the wireless network medium;
$b_{\max} \approx 109 \text{\caps{MB}}$, determined by the size of the largest flow in our trace data;
$t_{\max}=v_{\max}=600$, determined by our experimental methodology which slices traces into 10-minute scenarios;
and $v_{\min}=10^{-6}$, determined by the time resolution of our trace.
For lack of a better choice, we have simply chosen to use the same number of quantization bins as packet size for the unfixed dimension constants:
$d_a = d_b = d_v = 1500$.
Arguments can certainly be made for other choices, but this choice appears to work adequately in practice.

@@

From the matrices $\mat{Z}$ and $\mat{V}$ we can compute a variety of statistics about packet sizes and inter-packet intervals.
The row vectors in the matrices may represent individual flows or composites of many flows.
% about packet sizes, total packet count, inter-packet intervals, and total duration for each representation vector, which may represent a single actual flow, or the composite behavior of many flows.
% We begin by expressing the vectors for total bytes, total packets and total duration for a matrix of flow representations, with size and interval matrices $\mat{Z}$ and $\mat{V}$.
The computations require knowing the values that each size of interval matrix column represents.
% The computed duration is approximate because information is irretrievably lost in the quantization process.
The vectors of representative values for $\mat{Z}$ and $\mat{V}$ respectively are:
\begin{align}
\vec{\sigma} &= \tuple{1,2,3,\dots,d_z} \\
\vec{\tau}   &= \tuple{\Df_\ival\left(k-\smallfrac{1}{2}\right)}_{k=1}^{d_v}.
\end{align}
The packet count, total byte and total duration vectors are computed as
\begin{align}
\vec{\Sigma}_\pkts  &= \mat{Z}\ones = \mat{V}\ones \in \R^f \eqn{packet-count}\\
\vec{\Sigma}_\bytes &= \mat{Z}\trans{\vec{\sigma}} \in \R^f \eqn{total-bytes} \\
\vec{\Sigma}_\ivals &= \mat{V}\trans{\vec{\tau}} \in \R^f \eqn{total-duration}.
\end{align}
Here $\ones$ is a row vector of all ones.\onesnote
Packet count and total bytes values are exact;
the total durations are only approximate because information is irretrievably lost in the quantization process.
Using these vectors of totals, we may readily compute the average packet sizes and inter-packet intervals:
\begin{align}
\vec{\mean{z}} &=
  \vec{\Sigma}_\bytes \eldiv \vec{\Sigma}_\pkts
  \phantom{= \mat{V}\trans{\vec{\tau}} \eldiv \mat{V}\ones}
  \mathllap{= \mat{Z}\trans{\vec{\sigma}} \eldiv \mat{Z}\ones} \\
\vec{\mean{v}} &=
  \phantom{\vec{\Sigma}_\bytes \eldiv \vec{\Sigma}_\pkts}
  \mathllap{\vec{\Sigma_{\ivals}} \eldiv \vec{\Sigma_{\pkts}}}
  = \mat{V}\trans{\vec{\tau}} \eldiv \mat{V}\ones.
\end{align}
The symbol $\eldiv$ here represents element-wise division of vectors, as it does in Matlab.

@@

Given an instance of the general matrix model, how do we produce an actual workload from it?
Each row of the model is used to randomly generate the behavior of a single flow.
To do this, we use the ``dequantization'' functions listed in \Table{quantization}.
The dequantization procedure works as follows.
Let $\vec{f}$
% \begin{align}
% \vec{f}=\tuple{\vec{t},\vec{s},\vec{d},\vec{a},\vec{b},\vec{z},\vec{v}}
% \end{align}
be a row of the traffic matrix, and let $\prop$ be a property.
Write $\vec{f}_\prop$ for the corresponding component vector of $\vec{f}$.
Chose an index value $k\in\set{1,\dots,d_\prop}$, randomly, with weight given by the coefficients of $\vec{f}_\prop$.
Choose $u \in [0,1]$ uniformly at random.
Then $x = k-u \in [0,d]$, with probability of lying in the interval $[i-1,i]$ proportional to the coefficient of $\e_i$ in $\vec{f}_\prop$.
The dequantization function for $\prop$, namely $\Df_\prop$, will map the value $x$ back into the appropriate range of property values for $\prop$.
